Extended Data

### 2022-02-03 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
More rewording. @teixeirak i deleted one of the sentences around line 164. feel free to change back if you disagree

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures. 
In fact, we are aware of only a few studies that have documented stem-growth timing of temperate deciduous trees over multiple years[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022]. <!-- Could delete this sentence? I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
**However, the climate sensitivity of stem growth timing in temperate deciduous trees and its link to annual growth has seen a recent surge in interest within published literature[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; zweifel_why_2021]** <!-- revise this statement. Etzold et al kind of do this.-->

Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
To test whether warmer springs extend the period of stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
To test whether spring temperatures consistently increased annual growth, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.

@@

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperature window (CTW). 
CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem porosity-site combination (`r EDF_climwin`).
The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on annual ring-width (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to the growing season would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018; @etzold_number_2022].
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

>>>>

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures. 
This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021].

Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

Here, we evaluate how early spring temperatures affect peak stem growth timing, growth rates, and annual growth of temperate deciduous trees across eastern North America. 
To test whether warmer springs extend the period of peak stem growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
To test whether spring temperatures consistently increased annual growth, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.

@@

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW). 
CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem porosity-site combination (`r EDF_climwin`).
The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to the growing season would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018]. This decoupling between growing season length and annual growth mirrors results from another recent study by Etzold et al[@etzold_number_2022] which found that vapor pressure deficit plays a key role in limiting stem growth *- other studies which found VPD limits growth-*.
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.
